\vspace{-4em} \LARGE Hamiltonian Privilege

### 2023-05-25 Articles/2020-ClassicalMechanicsPhil/Submissions/ERKE/classical-phil-article-Erkenntnis.tex
Update classical-phil-article-Erkenntnis.tex
Added Acknowledgments and moved mention of AoP project there. Deleted all comments, so as to send source file to the editorial office.

\documentclass[12pt, twoside]{article}
\usepackage[a4paper]{geometry}
\geometry{top=4cm, bottom=4cm, left=2.5cm, right=2.5cm} 
%\geometry{top=4cm, bottom=4cm, left=3cm, right=4cm} %close to actual philsci margins

%\usepackage{endnotes}
%\let\footnote=\endnote %Erkenntnis requires end-notes, but recent papers don't seem to follow this in the pdf!
 % %To call the endnotes, use command \theendnotes before calling reference page

@@

\iffalse
\titleformat{\section}[runin]% runin puts it in the same paragraph
        {\normalsize\bfseries}% formatting commands to apply to the whole heading
        {\thesection.}% the label and number
        {0.5em}% space between label/number and section title
        {}% formatting commands applied just to section title
        [.]% punctuation or other commands following section title

\titlespacing*{\section}{0pt}{2.5mm}{2.5mm}%spacing from left margin, spacing above header, spacing on right of header
\fi

% headers
\usepackage{fancyhdr} 
\pagestyle{fancy}
\fancyhead{}
\fancyhead[R]{\thepage} %page number placement
\fancyhead[C]{\textsc{Hamiltonian Privilege}}
%\fancyhead[EC]{\textsc{AUTHOR NAMES}}%Even page header 
\fancyfoot[L,R,C]{}
\renewcommand{\headrulewidth}{0pt}
%\renewcommand{\footrulewidth}{0pt}
%\renewcommand*\footnoterule{} % removes footnote line

@@

@Article{Butterfieldpoints,
  author  = {Butterfield, Jeremy N.},
  journal = {British Journal for Philosophy of Science},
  title   = {Against Pointillisme about Mechanics},
  year    = {2006},
  pages   = {709--753},
  volume  = {57},
}

@@

@book{aop-book,
	author    = {Gabriele Carcassi and Christine A. Aidala},
	title     = {Assumptions of Physics},
	publisher = {Michigan Publishing},
	year      = 2021,
	isbn      = {978-1-60785-706-7},
	doi       = {10.3998/mpub.12204707},
	url       = {https://doi.org/10.3998/mpub.12204707}
}

@@

% % % Biblatex commands for Citations and References:
\usepackage[backend=bibtex, authordate, url=false, doi=false, eprint=false]{biblatex-chicago}
%\addbibresource{bibliography.bib}
\addbibresource{references.bib} %call the bib file generated above

% clears un-needed fields from reference items
\AtEveryBibitem{%
  \clearfield{note}%
  \clearfield{series}%
  \clearfield{pagetotal}%
   \clearfield{chapter}%
  \clearfield{number}%for convenience/consistency
}

@@

% Hyperref set-up & metadata
\hypersetup{
	pdftitle={Hamiltonian Privilege},
	pdfauthor={},
	pdfsubject={},
	pdfkeywords={},
	pdfcreator={XeLaTeX},
	pdfproducer={XeLaTeX},
	pdftoolbar=false,	% show Acrobat’s toolbar?
	pdfmenubar=true,	% show Acrobat’s menu?
	pdffitwindow=false,	% window fit to page when opened
	pdfstartview={FitH},	% fits the width of the page to the window
	unicode=true,		% non-Latin characters in Acrobat’s bookmarks
	pdfnewwindow=true,	% links in new window
	colorlinks=true,	% false: boxed links; true: colored links
	linkcolor=blue,	% color of internal links
	citecolor=black,	% color of links to bibliography
	filecolor=black,	% color of file links
	urlcolor=black		% color of external links
}

@@

\title{\vspace{-4em} \LARGE Hamiltonian Privilege}			
%\author{\Large \textsuperscript{*\textdagger}}
\date{\vspace{-7.9ex} \today}
\maketitle
%
\vspace{-8mm}

%\noindent\rule{\textwidth}{1.25pt}
\bs
\begin{flushleft}{\large \textbf{Abstract}}\end{flushleft}

\noindent {\normalsize We argue that Hamiltonian mechanics is more fundamental than Lagrangian mechanics. Our argument provides a non-metaphysical strategy for privileging one formulation of a theory over another: \textit{ceteris paribus}, a more general formulation is more fundamental. We illustrate this criterion through a novel interpretation of classical mechanics, based on three physical conditions. Two of these conditions suffice for recovering Hamiltonian mechanics. A third condition is necessary for Lagrangian mechanics. Hence, Lagrangian systems are a proper subset of Hamiltonian systems. Finally, we provide a geometric interpretation of the principle of stationary action and rebut arguments for privileging Lagrangian mechanics. }
\es
%\noindent\rule{\textwidth}{1.25pt}

%\vspace{2mm}

@@

Section~\ref{privileging} previews our argument for privileging Hamiltonian mechanics. We introduce three physical conditions that define classical systems while distinguishing Hamiltonian from Lagrangian mechanics. Section~\ref{fundamentality} explains how our position differs from North's. Specifically, our argument does not require metaphysical commitments to fundamental structure or perfectly natural properties. Section~\ref{derivation} describes in detail the derivation of Hamiltonian mechanics for systems that satisfy two physical conditions: infinitesimal reducibility and deterministic/reversible evolution. Our derivation shows that even non-hyper-regular Hamiltonian systems are physically interpretable. We then show that Lagrangian systems must satisfy an additional physical condition, which we call motion equivalence. This completes our primary argument for privileging Hamiltonian mechanics. Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle that applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics.\footnote{This paper forms part of the ongoing \textit{Assumptions of Physics} project \parencites[]{aop-book}. This project aims to both (i) isolate the physical assumptions required for different frameworks in mathematical physics and (ii) consider their philosophical implications.}

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must provide a criterion for assessing whether one framework is more fundamental than another. Of various notions of relative fundamentality, we focus on an entirely non-mysterious one. Given two formalisms that capture a class of systems, if one is logically more general, then it is more fundamental.\footnote{Hunt \parencites*[Ch. 5]{Huntthesis} provides an expressivist conception of fundamentality, based on an attitude of \textit{being for privileging}. In terms of this conception, adopting our specific criterion here expresses an attitude of being for privileging frameworks that are logically more general. Hunt \parencites*[816]{HuntPSA} provides a separate reason for privileging Hamiltonian mechanics: in systems with an ignorable coordinate, Hamiltonian mechanics makes manifest conservation of momentum.} This aspect of fundamentality is familiar from mathematics. A more fundamental framework encompasses a less fundamental one as a special case. It is in this sense that the generalized Stokes' theorem is more fundamental than special cases such as the fundamental theorem of calculus and Green's theorem.\footnote{One might worry that our criterion is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where ``aboutness'' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality must address. It is not a special problem facing our criterion.}  This aspect of fundamentality is distinct from aspects associated with differences in energy scale, such as how the Standard Model of particle physics is more fundamental than classical mechanics. We do not mean that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in this sense. As Gilton notes, ``many philosophers of science have moved away from thinking of fundamentality in science in terms of the very small toward thinking of it in terms of comprehensiveness, or breadth of applicability'' \parencites*[637]{Gilton2021}.
% % Removed: Similarly, an axiomatization that includes more axioms is less fundamental than one which drops these axioms while maintaining the same expressive power.
%Worry: what if  both axiom systems cover the same class of systems? then seemingly no difference in generality? 
%we must provide a criterion for one framework being more or less fundamental than another
%\footnote{In Section~\ref{density}, we embrace a related non-metaphysical notion of fundamentality that pertains to physical entities rather than theoretical frameworks.}

To apply our criterion for relative fundamentality, we must show that Lagrangian mechanics is less general than Hamiltonian mechanics. We do this in Section~\ref{derivation} by recovering the mathematical formalisms of both Hamiltonian and Lagrangian mechanics from three physical conditions. Our derivation illuminates the relationship between Hamiltonian and Lagrangian mechanics. As North \parencites*[]{North} argues, there is a sense in which Hamiltonian mechanics is more general than Lagrangian mechanics. Barrett \parencites*[]{Barrett2} makes this sense more precise by showing that the symplectic manifold formulation of Hamiltonian mechanics is categorically inequivalent to Lagrangian mechanics formulated on a tangent bundle. Still, one might wonder why it is that Hamiltonian mechanics can be formulated in a more general framework than Lagrangian mechanics. Additionally, one might wonder whether Lagrangian mechanics could be meaningfully formulated in an analogously general manner. Our derivation helps answer these questions: the physical principles that undergird the symplectic manifold formulation of Hamiltonian mechanics are simply more general than the principles required for Lagrangian mechanics.\footnote{We thank a referee for suggesting this helpful way of framing the significance of our derivation.} In this section, we describe how our argument works in principle. Section~\ref{fundamentality} demonstrates some philosophical advantages of our criterion for relative fundamentality.
%We will, however in Section 5 give a criterion for an entity to be physically fundamental
%These bolster our claim that this is an appropriate sense of fundamentality for philosophers to focus on.

% From Healey 2007, (weak) separability says that ``any physical process occupying space-time region R supervenes upon an assignment of qualitative intrinsic physical properties at space-time points in R (and/or in arbitrarily small neighborhoods of those points)'' [46]
% Non-separability: ``some physical process occupying a region R of space-time is not supervenient upon an assignment of qualitative intrinsic physical properties at space-time points in R'' [124]
% Strong non-separability: ``some physical process occupying a region R of space-time is not supervenient upon an assignment of qualitative intrinsic physical properties at points of R and/or in arbitrarily small neighborhoods of those points'' [124].

@@

Of course, different definitions of ``Hamiltonian'' and ``Lagrangian'' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems also satisfy the motion equivalence condition. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they lie in the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems. For a Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the details of our argument in Section~\ref{derivation}, we describe how we avoid problems that arise in North's privileging of Hamiltonian mechanics. 
%, although Curiel seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.

@@

As indicated at the start of Section~\ref{privileging}, we are interested in a non-metaphysical aspect of fundamentality. At least some differences in fundamentality are differences in generality: other things equal, a theoretical framework is more fundamental if it requires fewer conditions to hold. This aspect of fundamentality leads to a correspondingly non-metaphysical notion of structure. Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of comparative structure, our criterion has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which scientific anti-realists, pragmatists, and even some scientific realists view with suspicion. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure. 
% Could cite Woodward, or Cohen/Callender for skepticism about perfectly natural properties. Includes naturalistically-minded philosophers
%provided a theoretical framework captures the relevant behavior, it is more fundamental if it requires fewer conditions to hold.

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids these problems. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the motion equivalence condition. Motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general. 
%Section~\ref{density} provides a further argument for viewing Hamiltonian mechanics as more fundamental. We argue that (i) invariant phase space densities are physically fundamental and (ii) Hamiltonian mechanics provides the proper setting for representing these densities.
%is this the actual phase space density vs. density for all possible trajectories?

@@

Figure~\ref{diagram} summarizes the results of this derivation. The left column lists the three central physical conditions. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal/mathematical structure depicted in the right column, including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby elucidate the physical significance of these formalisms.
%details of this figure will become clear as we go along.

@@

%Like many arguments in physics, our derivation falls short of some norms on mathematical rigor. 
%optional to do: give a gloss on notion of regulative assumption (Kant and Peirce)

@@

First, we will show that a classical phase space arises if a system satisfies the following physical condition:
%\newpage % % %%remember to take out!

@@

\noindent
This assumption characterizes classical systems as those whose infinitesimal parts completely specify the state of the system. We define a \textit{classical particle} as one of these infinitesimal parts, namely the limit of recursive subdivision.\footnote{This definition of a classical particle differs from viewing point-particles as extensionless \parencites[]{Butterfieldpoints} or zero-dimensional \parencites[]{Wilson}.} Ultimately, this condition entails that we can describe the state of the whole system as a statistical distribution over the state space of its parts. If we model the state space of the parts using a manifold, the distribution must be characterized by a density that depends on the state and not on the coordinates. From this condition, we show that the manifold must be symplectic. Readers satisfied with this overview can proceed to Section~\ref{deterministic}.
% % Could add: readers who are satisfied with this overview can proceed to the next subsection
% % % For further discussion, see \textcites{shpmp}. [omit this for review].

Let $\mathcal{C}$ be the state space for the whole system, so that any $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point $s \in \mathcal{S}$ describes a possible state of a classical particle. Then for each system state $c \in \mathcal{C}$, there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state from $\mathcal{S}$.\footnote{For further details, see \textcites[]{AoPPhy1} and \textcites{shpmp}.}
%ref 1: worth clarifying relationship between particle statespace S and system state space C. Is there an S_i for each i-th particle? And explain why we are not focusing on the statespace of the many particle system, e.g. a cartesian product of state spaces for each particle

%If $\mathcal{S}$ is a discrete space with $U \subseteq \mathcal{S}$, we define $\mu(U) = \sum\limits_{s \in U} \rho_c(s)$ so that $\mu(\mathcal{S})$ gives us the size of the system and $\mu(U)/\mu(\mathcal{S})$ gives us the fraction of the particles that can be found in the region $U$.\footnote{The discrete case is far less interesting because discrete quantities cannot change continuously over time, meaning that they must be constant. We discuss this elsewhere.}
% % % For details see \textcites[]{AoPPhy1}. [omit this for review]}
% % XXX could try to find justification/derivation for the state space being a manifold in the Kevin Kelly book, or could cite the math paper with Greenfield; proposition 6.2. Try to include a better gloss/motivation for this claim (basic idea: if time is continuous, trajectories in state space must be continuous, local structure must be compatible). Curiel paper might also have a brief justification/motivation, if I recall

When the system is characterized by continuous quantities, $\mathcal{S}$ can be taken to be a manifold.\footnote{Note that the use of manifolds constitutes a choice of how to represent the physics. One could make another choice, for example, by representing the state variables not as local coordinates but as an algebra of operators that act on the state. In this case, we conjecture that a similar argument would demonstrate an analogue of a symplectic structure for this alternative mathematical representation.} For $U \subseteq \mathcal{S}$, we define $\mu(U) := \int_U \rho_c(s) d\mathcal{S}$ to measure the number of particles in region $U$. \textit{State variables} $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called ``coordinates'' and $\xi$ is a coordinate map. To avoid confusion, we will call the coordinates of the state space $\mathcal{S}$ ``state variables.'' As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument. It allows us to compare and track the amount of matter in different possible states.
%The use of manifolds is, technically, a choice of how the physics is represented

@@

We will now show that in order for the density $\rho_{c, \xi}$ to have these transformation properties, the particle phase space $\mathcal{S}$ must be at least a symplectic manifold. In other words, given that i) we assume we can model a classical system by a density over its infinitesimal states and ii) these transformation properties are required for the density to play this role, it follows that the phase space must have a symplectic manifold structure. Our next assumption involves the relationship between state variables and measurement units. 
%Since these transformation properties are required to use a density to model a classical mechanical system, it follows that having the structure of a symplectic manifold is likewise required.  
%From the preceding two equations, we see that $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.
% $d \xi^a$ and $d \hat{\xi}^b$ are the

@@

%on needing a two-form to measure areas, could possibly cite Strang 1993, ch. 5 or Mac Lane & Birkhoff (1999, Theorem IX.2.2). 
%note: an infinitesimal area counts the number of possible states in a region of phase space, as given by the two state variables that compose the degree of freedom

@@

%Moreover, the units we choose for $\omega$ are the units we choose to count states over an independent degree of freedom. This agrees with the practice in statistical mechanics.

@@

%Section~\ref{density} describes further interpretive consequences of this connection between invariant densities and symplectic structure.

@@

%Since, the only requirement that we have imposed on the Hamiltonian is that the equations yield a unique solution, our derivation applies to Hamiltonians that are not hyper-regular. }

% since we have not imposed hyper-regularity,   }
% [(ref 1): expand on how our argument relates to standard proofs of Liouville's theorem. Is it fair to say that we are privileging Hamiltonian mechanics because it forms the basis for a proof of Liouville's theorem?]

@@

% Somewhere below, emphasize that you have a unique trajectory only if the action is a minimum or maximum (if it is just stationary but not a minimum or maximum, then you do not have a unique trajectory)

@@

%\newpage

%\theendnotes

%\clearpage

>>>>

%% To compile bibliography, run Bibtex, which will generate a file `references.bib' which is called in the References section. 
%The file calls one diagram, labeled `Diagram.png'

\documentclass[12pt, twoside]{article}
\usepackage[a4paper]{geometry}
\geometry{top=4cm, bottom=4cm, left=2.5cm, right=2.5cm}

@@

\iffalse
\titleformat{\subsubsection}[runin]% for acknowledgements 
{\normalsize\bfseries}%
 {\thesection.}% the label and number
        {0.5em}% space between label/number and section title
        {}% formatting commands applied just to section title
        [.]% punctuation or other commands following section title

\fi

% headers
\usepackage{fancyhdr} 
\pagestyle{fancy}
\fancyhead{}
\fancyhead[R]{\thepage} %page number placement
%\fancyhead[C]{\textsc{Hamiltonian Privilege}}
\fancyhead[OL]{\small{Hamiltonian Privilege}}
\fancyhead[ER]{\small{Hunt, Carcassi, and Aidala}}%Even page header 
\fancyfoot[L,R,C]{}
\renewcommand{\headrulewidth}{0pt}
%\renewcommand{\footrulewidth}{0pt}
%\renewcommand*\footnoterule{} % removes footnote line

@@

@Article{Butterfieldpoints,
  author  = {Butterfield, Jeremy N.},
  journal = {The British Journal for the Philosophy of Science},
  title   = {Against Pointillisme about Mechanics},
  year    = {2006},
  pages   = {709--753},
  volume  = {57},
}

@@

@book{aop-book,
	author    = {Gabriele Carcassi and Christine A. Aidala},
	title     = {Assumptions of Physics},
	publisher = {Michigan Publishing},
	year      = 2021,
	doi       = {10.3998/mpub.12204707},
	url       = {https://doi.org/10.3998/mpub.12204707}
}

@@

% % % Biblatex commands for Citations and References:
\usepackage[backend=bibtex, authordate, url=false, doi=false, eprint=false]{biblatex-chicago}

\addbibresource{references.bib} %call the bib file generated above

% clears un-needed fields from reference items
\AtEveryBibitem{%
  \clearfield{note}%
  \clearfield{series}%
  \clearfield{pagetotal}%
   \clearfield{chapter}%
  \clearfield{number}%for convenience/consistency
  \clearfield{isbn}%
}

@@

% Hyperref set-up & metadata
\hypersetup{
	pdftitle={Hamiltonian Privilege},
	pdfauthor={Hunt, Carcassi, and Aidala},
	pdfsubject={},
	pdfkeywords={},
	pdfcreator={XeLaTeX},
	pdfproducer={XeLaTeX},
	pdftoolbar=false,	% show Acrobat’s toolbar?
	pdfmenubar=true,	% show Acrobat’s menu?
	pdffitwindow=false,	% window fit to page when opened
	pdfstartview={FitH},	% fits the width of the page to the window
	unicode=true,		% non-Latin characters in Acrobat’s bookmarks
	pdfnewwindow=true,	% links in new window
	colorlinks=true,	% false: boxed links; true: colored links
	linkcolor=blue,	% color of internal links
	citecolor=black,	% color of links to bibliography
	filecolor=black,	% color of file links
	urlcolor=black		% color of external links
}

@@

\title{\vspace{-4em} \LARGE Hamiltonian Privilege}			
%\author{\Large \textsuperscript{*\textdagger}}
\date{\vspace{-7.9ex} Josh Hunt, Gabriele Carcassi, and Christine Aidala \\ \today}
\maketitle
%
\vspace{-8mm}

\bs
\begin{flushleft}{\large \textbf{Abstract}}\end{flushleft}

\noindent {\normalsize We argue that Hamiltonian mechanics is more fundamental than Lagrangian mechanics. Our argument provides a non-metaphysical strategy for privileging one formulation of a theory over another: \textit{ceteris paribus}, a more general formulation is more fundamental. We illustrate this criterion through a novel interpretation of classical mechanics, based on three physical conditions. Two of these conditions suffice for recovering Hamiltonian mechanics. A third condition is necessary for Lagrangian mechanics. Hence, Lagrangian systems are a proper subset of Hamiltonian systems. Finally, we provide a geometric interpretation of the principle of stationary action and rebut arguments for privileging Lagrangian mechanics. }
\es

@@

Section~\ref{privileging} previews our argument for privileging Hamiltonian mechanics. We introduce three physical conditions that define classical systems while distinguishing Hamiltonian from Lagrangian mechanics. Section~\ref{fundamentality} explains how our position differs from North's. Specifically, our argument does not require metaphysical commitments to fundamental structure or perfectly natural properties. Section~\ref{derivation} describes in detail the derivation of Hamiltonian mechanics for systems that satisfy two physical conditions: infinitesimal reducibility and deterministic/reversible evolution. Our derivation shows that even non-hyper-regular Hamiltonian systems are physically interpretable. We then show that Lagrangian systems must satisfy an additional physical condition, which we call motion equivalence. This completes our primary argument for privileging Hamiltonian mechanics. Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle that applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics.

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must provide a criterion for assessing whether one framework is more fundamental than another. Of various notions of relative fundamentality, we focus on an entirely non-mysterious one. Given two formalisms that capture a class of systems, if one is logically more general, then it is more fundamental.\footnote{Hunt \parencites*[Ch. 6]{Huntthesis} provides an expressivist conception of fundamentality, based on an attitude of \textit{being for privileging}. In terms of this conception, adopting our specific criterion here expresses an attitude of being for privileging frameworks that are logically more general. Hunt \parencites*[816]{HuntPSA} considers a separate reason for privileging Hamiltonian mechanics: in systems with an ignorable coordinate, Hamiltonian mechanics makes manifest conservation of momentum.} This aspect of fundamentality is familiar from mathematics. A more fundamental framework encompasses a less fundamental one as a special case. It is in this sense that the generalized Stokes' theorem is more fundamental than special cases such as the fundamental theorem of calculus and Green's theorem.\footnote{One might worry that our criterion is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where ``aboutness'' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality must address. It is not a special problem facing our criterion.}  This aspect of fundamentality is distinct from aspects associated with differences in energy scale, such as how the Standard Model of particle physics is more fundamental than classical mechanics. We do not mean that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in this sense. As Gilton notes, ``many philosophers of science have moved away from thinking of fundamentality in science in terms of the very small toward thinking of it in terms of comprehensiveness, or breadth of applicability'' \parencites*[637]{Gilton2021}.

To apply our criterion for relative fundamentality, we must show that Lagrangian mechanics is less general than Hamiltonian mechanics. We do this in Section~\ref{derivation} by recovering the mathematical formalisms of both Hamiltonian and Lagrangian mechanics from three physical conditions. Our derivation illuminates the relationship between Hamiltonian and Lagrangian mechanics. As North \parencites*[]{North} argues, there is a sense in which Hamiltonian mechanics is more general than Lagrangian mechanics. Barrett \parencites*[]{Barrett2} makes this sense more precise by showing that the symplectic manifold formulation of Hamiltonian mechanics is categorically inequivalent to Lagrangian mechanics formulated on a tangent bundle. Still, one might wonder why it is that Hamiltonian mechanics can be formulated in a more general framework than Lagrangian mechanics. Additionally, one might wonder whether Lagrangian mechanics could be meaningfully formulated in an analogously general manner. Our derivation helps answer these questions: the physical principles that undergird the symplectic manifold formulation of Hamiltonian mechanics are simply more general than the principles required for Lagrangian mechanics.\footnote{We thank a reviewer for suggesting this helpful way of framing the significance of our derivation.} In this section, we describe how our argument works in principle. Section~\ref{fundamentality} demonstrates some philosophical advantages of our criterion for relative fundamentality.

@@

Of course, different definitions of ``Hamiltonian'' and ``Lagrangian'' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems also satisfy the motion equivalence condition. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they lie in the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems. For a Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the details of our argument in Section~\ref{derivation}, we describe how we avoid problems that arise in North's privileging of Hamiltonian mechanics.

@@

As indicated at the start of Section~\ref{privileging}, we are interested in a non-metaphysical aspect of fundamentality. At least some differences in fundamentality are differences in generality: other things equal, a theoretical framework is more fundamental if it requires fewer conditions to hold. This aspect of fundamentality leads to a correspondingly non-metaphysical notion of structure. Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of comparative structure, our criterion has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which scientific anti-realists, pragmatists, and even some scientific realists view with suspicion. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids these problems. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the motion equivalence condition. Motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general.

@@

Figure~\ref{diagram} summarizes the results of this derivation. The left column lists the three central physical conditions. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal/mathematical structure depicted in the right column, including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby elucidate the physical significance of these formalisms.

@@

@@

First, we will show that a classical phase space arises if a system satisfies the following physical condition:

@@

\noindent
This assumption characterizes classical systems as those whose infinitesimal parts completely specify the state of the system. We define a \textit{classical particle} as one of these infinitesimal parts, namely the limit of recursive subdivision.\footnote{This definition of a classical particle differs from viewing point-particles as extensionless \parencites[]{Butterfieldpoints} or zero-dimensional \parencites[]{Wilson}.} Ultimately, this condition entails that we can describe the state of the whole system as a statistical distribution over the state space of its parts. If we model the state space of the parts using a manifold, the distribution must be characterized by a density that depends on the state and not on the coordinates. From this condition, we show that the manifold must be symplectic. Readers satisfied with this overview can proceed to Section~\ref{deterministic}.

Let $\mathcal{C}$ be the state space for the whole system, so that any $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point $s \in \mathcal{S}$ describes a possible state of a classical particle. Then for each system state $c \in \mathcal{C}$, there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state from $\mathcal{S}$.\footnote{For further details, see \textcites[]{AoPPhy1} and \textcites{shpmp}.}

When the system is characterized by continuous quantities, $\mathcal{S}$ can be taken to be a manifold.\footnote{Note that the use of manifolds constitutes a choice of how to represent the physics. One could make another choice, for example, by representing the state variables not as local coordinates but as an algebra of operators that act on the state. In this case, we conjecture that a similar argument would demonstrate an analogue of a symplectic structure for this alternative mathematical representation.} For $U \subseteq \mathcal{S}$, we define $\mu(U) := \int_U \rho_c(s) d\mathcal{S}$ to measure the number of particles in region $U$. \textit{State variables} $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called ``coordinates'' and $\xi$ is a coordinate map. To avoid confusion, we will call the coordinates of the state space $\mathcal{S}$ ``state variables.'' As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument. It allows us to compare and track the amount of matter in different possible states.

@@

We will now show that in order for the density $\rho_{c, \xi}$ to have these transformation properties, the particle phase space $\mathcal{S}$ must be at least a symplectic manifold. In other words, given that i) we assume we can model a classical system by a density over its infinitesimal states and ii) these transformation properties are required for the density to play this role, it follows that the phase space must have a symplectic manifold structure. Our next assumption involves the relationship between state variables and measurement units.

@@

@@

@@

@@

@@

@@

\subsubsection*{Acknowledgements}
